---
title: Creating a cartogram using Mark Newman's algorithm based on diffusion maps 
author: Philipp Müller
output: rmarkdown::md_document
---

This file handles the generation of the cartogram according to Mark Newman's [algorithm](http://www-personal.umich.edu/~mejn/cart/) and the corresponding R implementation.

# Prerequisites
First you have to load the required packages R packages and all shape data of the election districts.

```{r "init"}
library( devtools ) # To install the package from Github
library( ggplot2 ) # For plotting the results
library( dplyr ) # For reshaping and selecting the election results
library( rgdal ) # For importing the shape files
library( maptools ) # and their handling

## Loading the election results.
load( "../data/bundeswahlleiter/current_election.RData" )

## Import the shape files
election.districts <- readOGR( dsn = "../data/bundeswahlleiter/.",
                              layer = "Geometrie_Wahlkreise_19DBT_geo",
                              p4s = NULL )

## Converting the spatial object into a data.frame better suited
## for the 'ggplot2' package.
election.districts.df <- fortify( election.districts )
```

### Installing the libfftw3 package (Linux users)
Before we can install this package we need to do a little bit of
configuration. The package itself links against the *fftw.h* header
of the **libfftw3** package. It's most probably already installed on your
system. But for some reason the installation fails since the package
is ill-configured. Therefore we have to download its source code,
reconfigure it using the *--enable-shared* option, and compile and
install it ourselves.
Don't be afraid. The worst thing that could happen here is for your
R package to break after a dist. upgrade of your system. But be sure
to compile and install the libfftw3 package in a folder you won't
touch or remove in the future! Else your system can not access its
functions anymore.

Depending on your system a different version of the libfftw3 package
might be installed.
```{bash}
apt search fftw3 | grep libfftw3
```
The package we are looking for is usually the first one appearing. It
should contain the basename *libfftw3* and maybe an additional number
attached using a dash, but definitely no characters like 'dev', 'dbg'
etc!

Now let's download its source code, reconfigure, recompile, and re-
install it.
```{bash}
# Download the package source
sudo apt source libfftw3-3
# Reconfigure the package with the fftw.h available to link against
sudo ./configure --enable-shared
# Compile the package
sudo make
# Installation
sudo make install
```

### Install the Rcartogram package
Now we can head over to R again to install the Rcartogram
Since it's not on CRAN we will use the most recent Github version.

```{r "init2", dependson="init"}
install_github( "omegahat/Rcartogram" )
require( Rcartogram )
```
# Density generation
The package's code is a port to Newman's [original C version](http://www-personal.umich.edu/~mejn/cart/doc/) 
of the algorithm. 
The basic ingredient we need here is a density map on a regular
grid containing the election results per election district.
This map we will afterwards embed in a 'sea' of the overall
average value of quite generous size.

We will extract the overall range of the polygons of the election districts
and make the size of the grid explicit.

```{r "grid-range", dependson="init2"}
election.districts.range <- election.districts@bbox
number.of.grid.points.y <- 1024
number.of.grid.points.x <- round( number.of.grid.points.y/ 2 )
```

Since I had the impression that bigger ranges result in less distortion
we span an uniform grid of points covering the very range of the election districts.
We will use the SpatialPoints object instead of a data.frame
here, since the use of a spatial object defined in the **sp** package
is required by the **over** function we'll use in the next step. This function
checks if a supplied point falls within a polygon. The output of this function
we will use to associate our grid point with the number of votes for the election
districts they belong to.
```{r "grid-spanned", dependson="grid-range"}
density.grid.x <-  seq( election.districts.range[ 1, 1 ],
                       election.districts.range[ 1, 2 ],
                       length.out = number.of.grid.points.x )
density.grid.y <-  seq( election.districts.range[ 2, 1 ],
                       election.districts.range[ 2, 2 ],
                       length.out = number.of.grid.points.y )
density.grid <- SpatialPoints( expand.grid(
    density.grid.x, density.grid.y ),
    proj4string = election.districts@proj4string )

## Check which point falls within which polygon/election
## district.
density.point.in.polygon <- over( density.grid, election.districts )
## Collect the results in a data.frame
density.df <- data.frame(
    x = density.grid@coords[ , 1 ], y = density.grid@coords[ , 2 ],
    election.district.number =
      as.numeric( as.character(
          density.point.in.polygon$WKR_NR ) ) )
```

as.numeric( as.character( ... ) )? Well, without the
conversion to character the numerical conversion will
output wrong number in the right range resulting in a
wrong distribution of the election districts. Those are the most ***** bugs.

For this example let's plot the preliminary second votes for the leftist party again.
```{r "selected-party", dependson="grid-range"}
## This code chunk does not really depend on the last but
## on even earlier ones, but let's be simple and linear in here
select.party <- "DIE LINKE"
select.type.of.vote <- "Zweitstimmen"
select.confirmation.status <- "Vorläufig"
select.counts <- select(
    filter( current.election,
           party == select.party &
           type.of.vote == select.type.of.vote &
           confirmation.status == select.confirmation.status &
           county.number != 99 ),
    election.district.number, counts )
```
In addition we have to prepare the election district data object. We convert it into
a data.frame holding only the coordinates of the polygons' vertices, increment the
*id* key to represent the number of the election district, and use it to relate our
shapes to the number of votes we just filtered.

```{r "selected-districts", dependson="selected-party"}
election.districts.df <- fortify( election.districts )

## Convert the IDs of the polygons to the same key used to
## identify the election districts.
election.districts.df$election.district.number <-
  as.numeric( election.districts.df$id ) + 1

## Add the counts to all points of the corresponding polygons
election.districts.df <- left_join( election.districts.df,
                                   select.counts,
                                   by ="election.district.number")
``` 
In a similar manner we will now relate number of votes to the grid points
we got from the previous point in polygon test. In addition we will replace all missing
values for grid points belonging to no election district by the mean number of votes
for the leftist party.

```{r "density-df", dependson="selected-districts"}
## Assign the corresponding number of votes to the election
## district the individual points did fall into.
density.df <- left_join( density.df, select.counts,
                        by = "election.district.number" )

## Fill all NA in the number of votes with the mean value.
## This is what the authors called the 'sea'.
density.df$counts[ is.na( density.df$counts ) ] <-
  mean( density.df$counts, na.rm = TRUE )
```

And do a short visual check of the density map using the **ggplot2** package.
```{r, dependson="selected-districts"}
ggplot( data = density.df ) +
  geom_tile( aes( x = x, y = y, alpha = counts ),
            fill = '#df0404' ) +
coord_quickmap() + theme_minimal() + xlab( "" ) + ylab( "" ) +
  scale_alpha_continuous( guide = FALSE )
```
Alright, everything looks fine.

All that left to do is to convert the density into matrix.
```{r "density-matrix", dependson="density-df"}
density.matrix <- matrix( as.numeric( density.df$counts ),
                         nrow = number.of.grid.points.x,
                         ncol = number.of.grid.points.y )

## These intermediate step you can visualize using e.g
## image( density.grid.x, density.grid.y, density.matrix )
```

# Calculating the cartogram
### Using the Rcartogram package

The most easy and straight forward interface to Mark Newman's code
is of course the Rcartogram package.

We will use it the calculate the transformed density
```{r "cartogram-r-calc",dependson="density-matrix"}
density.cartogram <- cartogram( density.matrix )
```
and use the *predict()* function to obtain the polygons of the
election districts in the transformed density.
```{r "cartogram-r",dependson="cartogram-r-calc"}
cartogram.values.R <- predict( density.cartogram,
                            election.districts.df$long,
                            election.districts.df$lat )
```
### Using Newman's pure C code

As an alternative I provided the original C code of Newman
[version 1.2.2](http://www-personal.umich.edu/~mejn/cart/download/).
As a prerequisite we have to create an auxiliary folder, copy and
and compile the C files, and write both the density matrix and the
coordinates of the election districts to disk. After the
transformation we will delete this folder.
```{r "cartogram-c",dependson="cartogram-r"}
dir.create( "tmp" )
file.copy( c( "../src/main.c", "../src/cart.c", "../src/cart.h",
             "../src/interp.c" ), to = "tmp/" )
setwd( "tmp" )

## Compile the C source. For more detailed information see
## http://www-personal.umich.edu/~mejn/cart/doc/
system2( "gcc", args = "-O -o cart cart.c main.c -lfftw3 -lm" )
system2( "gcc", args = "-O -o interp interp.c" )

## Write the density matrix and the election districts to disk
write( t( density.matrix ), "densityMatrix.dat",
      ncolumns = number.of.grid.points.x )
positions.election.districts.matrix <- as.matrix(
    select( election.districts.df, long, lat ) )
## This is an incredibly inefficient way of reshaping the matrix
## but in order to process the positions as a stream it is necessary.
positions.election.districts <- Reduce(
    c, apply( positions.election.districts.matrix, 1, c ) )
write( positions.election.districts, file = "districts.dat",
      ncolumns = 2 )

## Calculate the transformed density
system2( "./cart",
        args = paste( number.of.grid.points.y,
                     number.of.grid.points.x,
                     "densityMatrix.dat densityTransformed.dat" ) )
## Transform the vertices of the election districts
system2( "cat",
        args = paste( "districts.dat | ./interp",
                     number.of.grid.points.y,
                     number.of.grid.points.x,
                     "densityTransformed.dat > districtsTransformed.dat" ) )

## Import the transformed election district data into R
cartogram.values.c <- read.table( "districtsTransformed.dat",
                                 header = FALSE, row.names = NULL )

## Cleanup
setwd( ".." )
unlink( "tmp" )
```

# Plotting the results

Almost done. Now we will combine the transformed coordinates, their ID for 
the association to a particular polygon, and the election results into data.frames.
Those we again combine into a final one better suited for the plotting procedure.
```{r "plot-prepare",dependson="cartogram-c"}
## Combining the results into a data.frame
cartogram.df.R <- data.frame( long = cartogram.values.R[[ 1 ]],
                           lat = cartogram.values.R[[ 2 ]],
                           group = election.districts.df$group,
                           counts = election.districts.df$counts )
## Combining the results into a data.frame
cartogram.df.c <- data.frame( long = cartogram.values.c[[ 1 ]],
                           lat = cartogram.values.c[[ 2 ]],
                           group = election.districts.df$group,
                           counts = election.districts.df$counts )

## and compare it to the default plot.
plot.total <- data.frame(
    long = c( election.districts.df$long, cartogram.df.R$long,
             cartogram.df.c$long),
    lat = c( election.districts.df$lat, cartogram.df.R$lat,
            cartogram.df.c$lat ),
    counts = c( election.districts.df$counts, rep( cartogram.df.R$counts, 2 ) ),
    group = c( election.districts.df$group, rep( cartogram.df.R$group, 2 ) ),
    type = factor( c( rep( "default", nrow( election.districts.df ) ),
                        rep( "Rcartogram", nrow( cartogram.df.R ) ),
                        rep( "cart c function", nrow( cartogram.df.R ) ) ),
                  levels = c( "default", "Rcartogram", "cart c function" ) ) )
```

Finally we plot the results.
```{r,dependson="plot-prepare"}
ggplot() + geom_polygon( data = plot.total,
                          aes( x = long, y = lat, group = group,
                              alpha = counts ),
                        colour = "black", fill = "#df0404" ) +
coord_quickmap() + theme_minimal() + xlab( "" ) + ylab( "" ) +
  scale_alpha_continuous( guide = FALSE ) +
  facet_wrap(~ type )
```
